Using VariantSpark to learn population structure from genomic profiles

### 2015-10-27 doc/publications/Recomb15/recomb15.tex
updates

\maketitle
%
\begin{abstract}
Using VariantSpark, we compare various machine learning algorithms on the 1092 individuals in the 1000 Genomes Project dataset.
Although each algorithm that we investigate can classify or cluster individuals from three populations with near perfect accuracy,
we find logistic regression to outperform the others. We observe a similar outcome on the entire dataset, with logistic regression
classifying the majority of the individuals correctly.

@@

\marginpar{I don't understand this. Can't you just ask a patient about their ancestry?}
Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP
information~\cite{Zheng2014}. Such clinical application are severely hampered for patients where accurate population
association is unknown due to a diverse migrational background. Determining the population association from genomic
information may provide vital insights for treatment and diagnostic in such cases.

@@

Using \kMeans{} we observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European
(EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR
individual form an independent cluster while the majority falls into the EUR cluster.  We quantified this by calculating
the \ARI\ (ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI of 0.84, with -1 being
random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an unsupervised method
rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise that a supervised
machine learning method may be capable of separating these subgroups of low genetic divergence.  Indeed, bigdata
genomics recently investigated the supervised method, deep learning, for clustering the 1000 Genomes data by
ethnicity~\cite{Ferguson}.  \marginpar{``bigdata genomics'' seems to be a domain name \url{http://bdgenomics.org/} --
  how should we cite this}

@@

To demonstrate \variantSpark's capability, we determine the cross-validation accuracy of the supervised multivariate RF
method and compare it against the simple supervised method LR as well as the unsupervised \kMeans{} in the first
section.  In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate
deep learning (DL) method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.

@@

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
rather than the number of samples, and a potential solution is to either limit the features through feature selection, or to simply use a subset of the dataset.
We therefore use only chromosome 22 (494,328 variants) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes Project dataset using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and calculate the Adjusted Rand Index (ARI) for use as a metric for model quality.
This metric allows us to compare the classification results with \kMeans{} clustering results.
We also report the mean error for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.

We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)). As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately,
with an average ARI over ten models of \(0.596 \pm 0.067\) (see table~\ref{RF}).
%Although capable of fitting a relatively accurate model (\(ARI=0.837\)), The average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it is outperformed by the LR, with an ARI of \(0.923 \pm 0.015\).
This indicates that while applying a supervised learning method results in higher accuracy, a multivariate approach may not be needed to accurately fit ethnicity.

@@

Although we see a slight improvement with logistic regression in regards to other algorithms, their results are still very similar.
However, there is not much room in the first place for logistic regression to demonstrate any potential increases in model accuracy.
We therefore proceed to apply our method to the entire 1000 Genomes Project phase 1 dataset, which we previously clustered with an ARI of 0.84.

@@

\subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
In this section we apply both the supervised LR and unsupervised \kMeans\ method on the full dataset of 14 populations and assess the 
classification accuracy into the 4 super-populations.
Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
This is an improvement on our previous result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.83\).

\begin{table}
\caption{Cross-validated results, mean accuracy (and standard error) on the 4 super-populations.}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
method  & ARI & error \\
LR  & 0.936 (0.014) & 0.029 (0.005) \\
\kMeans & 0.84 & -- \\
\noalign{\smallskip}
\hline
\end{tabular}
\end{center}
\end{table}

To determine where the model is underperforming, we also record the precision and recall for each of the groups.
We observe that for each model, the precision and recall are always 1.0 for the EAS super-population.
This demonstrates that each logistic regression model was able to classify every EAS individual correctly, while not misclassifying any other individuals as EAS.
For the AMR group, however, we consistently observed a relatively low precision (\(0.829 \pm{} 0.02 \)).
This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
Consequently, this leads to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
Although these misclassifications as AMR are relatively frequent, we did not observe any AFR individuals being classified as EUR, or {\it vice versa}, in any of the ten folds.
Also, only in one fold did we observe one AMR individual misclassified as another super-population (AFR).
These results demonstrate the diversity, or relative difference of some of the groups.

\begin{table}
\caption{Mean precision and recall (and standard error of mean) for each super-population across the ten logistic regression folds.}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
super-population  & precision & recall \\
AFR  & 0.997 (0.003) & 0.956 (0.013) \\
AMR & 0.829 (0.020) & 0.994 (0.005) \\
EAS  & 1 (0) & 1 (0) \\
EUR  & 1 (0) & 0.950 (0.015) \\
\noalign{\smallskip}
\hline
\end{tabular}
\end{center}
\end{table}

\subsubsection{\kMeans{} clustering is faster and consumes less memory than random forests}

Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests (see table~\ref{resources}).
This is likely due to RF being an ensemble method. For example, where logistic regression and \kMeans{} clustering build a single model,
RF is building an forest of decision trees.
When training on chromosome 22 (494,328 variants), resource requirements are kept to a minimum, with no more than 3GB memory required per executor (node)
(6GB for RF). Although we specify 80 executors, the same algorithm could successfully complete on a much smaller cluster, at the expense of time. 
However, when we train the algorithms on the entire genome, memory requirements increase to 24GB. This limited us to using 40 executors, due to memory availability.
The memory requirements also restricts a job of that size to clusters with at least 24GB available. This demonstrates that pre-processing the data to reduce its dimensionality will
be an essential step for future work, especially as datasets grow outwards.

@@

\section{Conclusion}
In this paper we systematically investigated the performance of supervised and unsupervised methods for determining the
ethnicity from whole genome variant data.  We found that a supervised method, logistic regression, outperforms the
unsupervised approach, \kMeans{} (0.936 vs 0.84).  We also found that ethnicity can be better predicted by the simple
logistic regression method compared to the 
%unsupervised \kMeans{} approach or the 
multivariate random forest or deep learning approach as reported by
Big Data Genomics using the ADAM framework (1 vs 0.96 and 0.976, respectively).  Furthermore, we observe a limitation within
Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1096 individuals
of the 1000 Genomes Project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their
70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression
outperforms all other methods it remains unable to accurately identify all individuals from EUR (lowest precision) or misclassify AMR (lowest recall) individuals. 
We therefore ague that the remaining mixture is potentially reflective of the migrational background, which may be an important observation for medical
applications that need accurate population association.

>>>>

\maketitle
%
\begin{abstract}
Using VariantSpark, we compare various machine learning algorithms on the 1000 Genomes Project dataset.
The algorithms we investigate classify or cluster the individuals from three populations with near perfect accuracy,
with logistic regression outperforming the others.
We proceed to investigate the algorithms on all four super-populations, where we also observe
logistic regression to outperform the rest.
However, even with the parallel nature of {\sc Spark}, resource requirements increase dramatically with the feature size,
indicating that feature selection will be a requirement for larger datasets.

@@

%\marginpar{I don't understand this. Can't you just ask a patient about their ancestry?}
Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP
information~\cite{Zheng2014}. Such clinical application are severely hampered for patients where accurate population
association is unknown due to a diverse migrational background. Determining the population association from genomic
information may provide vital insights for treatment and diagnostic in such cases.

@@

Using \kMeans{} we observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European
(EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR
individual form an independent cluster while the majority falls into the EUR cluster.  We quantified this by calculating
the \ARI\ (ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI of 0.84, with -1 being
random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an unsupervised method
rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise that a supervised
machine learning method may be capable of separating these subgroups of low genetic divergence.  Indeed, bigdata
genomics recently investigated the supervised method, deep learning, for clustering the 1000 Genomes data by
ethnicity~\cite{Ferguson}.
%\marginpar{``bigdata genomics'' seems to be a domain name \url{http://bdgenomics.org/} --  how should we cite this}

@@

To demonstrate \variantSpark's capability, we determine the cross-validation accuracy of the supervised multivariate RF
method and compare it against the simple supervised method LR as well as the unsupervised \kMeans{} in the first
section.  In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate
deep learning (DL) method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{} clustering.

@@

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and LR.  
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
rather than the number of samples, and a potential solution is to either limit the features through feature selection, or to simply use a subset of the dataset.
We therefore use only chromosome 22 (494,328 variants) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes Project dataset using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and calculate the Adjusted Rand Index (ARI) for use as a metric for model quality.
This metric allows us to compare the classification results with \kMeans{} clustering results.
We also report the mean error for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.

We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)). As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately,
with an average ARI over ten models of \(0.596 \pm 0.067\) (see table~\ref{RF}).
%Although capable of fitting a relatively accurate model (\(ARI=0.837\)), The average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it is outperformed by LR, with an ARI of \(0.923 \pm 0.015\).
This indicates that while applying a supervised learning method results in higher accuracy, a multivariate approach may not be needed to accurately fit ethnicity.

@@

Although we see this slight improvement with logistic regression in regards to other algorithms, their results are still very similar.
However, there is not much room in the first place for logistic regression to demonstrate any potential increases in model accuracy.
We therefore proceed to apply our method to the entire 1000 Genomes Project phase 1 dataset, which we previously clustered with an ARI of 0.84.

@@

\subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
In this section we apply both the supervised LR and unsupervised \kMeans\ method on the full dataset of 14 populations and assess the 
classification accuracy into the 4 super-populations.
Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
This is an improvement on our result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.84\) (see table~\ref{ALLCHR}).

\begin{table}
\caption{Cross-validated results, mean accuracy (and standard error of the mean) on the 4 super-populations.}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
method  & ARI & error \\
LR  & 0.936 (0.014) & 0.029 (0.005) \\
\kMeans & 0.84 & -- \\
\noalign{\smallskip}
\hline
\label{ALLCHR}
\end{tabular}
\end{center}
\end{table}

To determine where the model is underperforming, we also record the precision and recall for each of the groups (see table~\ref{PR}).
We observe that for each model, the precision and recall are always 1.0 for the EAS super-population.
This demonstrates that each logistic regression model was able to classify every EAS individual correctly, while not misclassifying any other individuals as EAS.
For the AMR group, however, we consistently observed a relatively low precision (\(0.829 \pm{} 0.02 \)).
This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
Consequently, this leads to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
Although these misclassifications as AMR are relatively frequent, we did not observe any AFR individuals being classified as EUR, or {\it vice versa}, in any of the ten folds.
Also, only in one fold did we observe one AMR individual misclassified as another super-population (AFR).
These results demonstrate the diversity, or relative difference of some of the groups.

\begin{table}
\caption{Mean precision and recall (and standard error of mean) for each super-population across the ten logistic regression folds.}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
super-population  & precision & recall \\
AFR  & 0.997 (0.003) & 0.956 (0.013) \\
AMR & 0.829 (0.020) & 0.994 (0.005) \\
EAS  & 1 (0) & 1 (0) \\
EUR  & 1 (0) & 0.950 (0.015) \\
\noalign{\smallskip}
\hline
\label{PR}
\end{tabular}
\end{center}
\end{table}

\subsubsection{\kMeans{} clustering is faster and consumes less memory than random forests and logistic regression}

Time and resource consumption vary greatly between the three algorithms, with the most resource intensive being random forests (see table~\ref{resources}).
This is likely due to RF being an ensemble method. For example, where logistic regression and \kMeans{} clustering build a single model,
RF is building an forest of decision trees. Although memory requirements for LR are identical to \kMeans{} clustering, LR requires more than twice as much time.

When training on chromosome 22 (494,328 variants), resource requirements are kept to a minimum, with no more than 3GB memory required per executor (node)
for \kMeans and LR, with 6GB required for RF. Although we specify 80 executors, the same algorithm could successfully complete on a much smaller cluster, albeit at the expense of time. 
However, when we train the algorithms on the entire genome, memory requirements increase to 24GB. This limited us to using 40 executors, due to memory availability, and would also restrict
a job of this size to clusters with at least 24GB available. This demonstrates that pre-processing the data to reduce its dimensionality will
be an essential step for future work, especially as datasets grow outwards.

@@

\section{Conclusion}
In this paper we systematically investigated the performance of supervised and unsupervised methods for determining the
ethnicity from whole genome variant data.  We found that a supervised method, logistic regression, outperforms the
unsupervised approach, \kMeans{} (0.936 vs 0.84).  We also found that ethnicity can be better predicted by the simple
logistic regression method compared to the 
%unsupervised \kMeans{} approach or the 
multivariate random forest or deep learning approach as reported by
Big Data Genomics using the ADAM framework (1 vs 0.96 and 0.976, respectively).  Furthermore, we observe a limitation within
Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to approximately 10 million features for the 1096 individuals
of the 1000 Genomes Project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their
70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression
outperforms all other methods it remains unable to accurately identify all individuals from EUR (lowest precision) or misclassify AMR (lowest recall) individuals. 
We therefore ague that the remaining mixture is potentially reflective of the migrational background, which may be an important observation for medical
applications that need accurate population association.
